\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes,amsmath}
\begin{document}

\date{}

\title{\Large \bf Real-Time Detection of Lateral Movement in Cloud VPC Networks via Graph-Based Analysis of Flow and Authentication Logs}

\author{
{\rm Ibrahim Pehlivan}\\
University of California, Los Angeles
\and
{\rm Wesley Gunawan}\\
University of California, Los Angeles
\and
{\rm Samyak Kakatur}\\
University of California, Los Angeles
}

\maketitle

\thispagestyle{empty}


\subsection*{Abstract}
\begin{itemize}
\item Problem: lateral movement in cloud VPCs is hard to detect because internal traffic is trusted by default and individual log sources provide incomplete visibility.
\item Approach: build a streaming graph from combined network flow logs and authentication logs, extract structural, temporal, and statistical features, and score nodes/edges for lateral movement likelihood.
\item Key question: does combining both log sources via graph analysis improve detection accuracy and latency compared to single-source baselines?
\item Setup: analysis of two public datasets: LANL-Dataset-2015 (58 days of cloud VPC network events from Los Alamos National Laboratory with red-team ground truth) and DAPT2020 (simulated APT attack scenarios with labeled attack stages).
\item Expected contributions: graph construction pipeline from multi-source logs, combined detection method outperforming single-source baselines, analysis of which graph features best separate attack traffic from normal operations.
\end{itemize}

\section{Introduction}

Cloud VPC networks present a complex security landscape where internal traffic flows freely between hosts by default. Lateral movement, the phase of an attack where an adversary pivots through a compromised network to reach high-value targets, is especially difficult to detect because the volume of benign connections dwarfs the few that matter. Existing intrusion detection tools were built for perimeter monitoring and struggle with the noise and dynamism of cloud VPC environments, where diverse user activity and constant connection churn mask attacker activity.

No single log source provides enough signal to reliably identify lateral movement. Network flow logs capture connections but cannot distinguish an attacker's session from an admin's; authentication logs record login events but are blind to reconnaissance via port scans or data exfiltration that never triggers a login. Prior detection methods have relied on either signature matching, which misses novel attack patterns, or anomaly detection on a single log source, which floods analysts with false positives on busy production networks. We propose combining both log types into a unified streaming graph where computers and identities are nodes, connections and authentication events are edges, and graph-based feature analysis can detect the structural and temporal patterns that distinguish lateral movement from routine operations. This paper presents our graph construction pipeline, the detection methodology, and an evaluation on the LANL-Dataset-2015 and DAPT2020 datasets against single-source baselines.

\section{Related Work}


\textbf{Where Related Work Should Go:} We decided to place it near the top, as we believe it's "essential for motivation". We want to clearly start off explaining existing strategies and implementations, and then explain their limitations. Then, we can go deep into what makes our approach different (graph-based detection from multiple log sources). This allows the reader to clearly separate our new ideas from what already exists, and justifies it due to the limitations.
\newline

Graph-based detection methods model relationships between system events and reason over paths or substructures. Hopper~\cite{hopper2021} builds a login graph from authentication events and uses path inference to flag lateral movement sequences. However, its reliance on authentication logs alone means it cannot detect movement that bypasses login entirely, such as exploitation of services or use of stolen session tokens. We extend this idea by incorporating network flow logs, capturing lateral movement paths that never produce an authentication event. Euler~\cite{euler2023} applies temporal link prediction to cloud VPC network graphs, treating lateral movement as a link prediction problem in an evolving graph. Its temporal modeling is a strong baseline for our own temporal graph features, though it operates on a single data source. POIROT~\cite{poirot2019} matches provenance graph templates against audit logs to detect known attack patterns. Template matching is effective when attack structures are known in advance but cannot catch novel variations; we use anomaly-based scoring instead and adapt the graph model to cloud VPC network logs rather than host-level audit data.

Multi-log correlation approaches combine evidence from separate data sources to improve detection coverage. HOLMES~\cite{holmes2019} correlates system audit logs across multiple stages of the kill chain, mapping observed behavior to MITRE ATT\&CK tactics and producing a real-time threat score. Its multi-stage correlation confirms that combining log sources can improve detection, and its kill-chain framing shaped our detection pipeline structure. However, HOLMES operates on host-level audit logs in traditional enterprise settings, whereas we target network flow logs and authentication events at cloud VPC network scale.

\section{Background and Problem Setting}
\begin{itemize}
\item Lateral movement definition: post-exploitation phase where attacker pivots through network to reach targets, sits between initial access and objectives (data exfiltration, privilege escalation)
\item Kill-chain context: reconnaissance, initial compromise, lateral movement, objective completion
\item Cloud VPC architecture: virtual private cloud with VMs, subnets, firewall rules; VPC flow logs record source/dest IP, ports, bytes, packets; cloud audit/auth logs record SSH attempts, service account usage, IAM events
\item Threat model: assumes one computer already compromised (e.g., via phishing or vulnerable service); attacker goals include reaching high-value targets, expanding access via credential reuse, exfiltrating data; attacker capabilities include port scanning, SMB lateral movement, credential stuffing, service exploitation; red-team ground truth provides labeled attack events
\item Why existing tools fall short: perimeter IDS sees only north-south traffic; host-based agents miss network context; SIEM correlation rules are brittle and require manual tuning; network flow analysis alone lacks authentication context
\end{itemize}

\section{Methods and Approach}

We ask two questions: does combining flow and authentication logs in a graph improve detection over single-source methods, and how does this approach compare to classical anomaly detection on tabular features?

\subsection{Design Rationale}

Our pipeline is built around how lateral movement actually works. Attackers rely on TCP-based services like SMB (port 445), SSH (port 22), RDP (port 3389), because these protocols give them reliable, ordered delivery for remote command execution and file transfer. During reconnaissance, a compromised host connects to many destinations in quick succession (high fan-out), then narrows to specific targets once a path opens up. Attacker timing also differs from normal users: lateral movement tends to involve deliberate, spaced-out connections rather than the rapid-fire patterns of routine traffic.

These patterns shape our design. We include protocol and port features in edge scoring because lateral movement is inherently service-specific. We compute fan-out and degree features at the node level to catch the scanning-to-targeting transition. We extract temporal features (inter-arrival times, burst scores, active duration) to separate deliberate attacker behavior from routine traffic.

\subsection{Graph Construction}

We model the network as a directed multigraph using the igraph library. Nodes are entities (computers and users); edges are interactions (authentication events and network flows). The graph is built incrementally by a \texttt{StreamingGraphBuilder} that reads log files line-by-line through time windows derived from the red-team ground truth. Events outside all windows are discarded immediately, keeping memory proportional to the active window set rather than the full dataset. The pipeline processes over 135 million events in approximately 4 GB of RAM.

For authentication events, the builder creates a computer-to-computer edge (source computer to destination computer) and optionally a user-to-user edge. Each edge stores authentication type, logon type, orientation, success status, and timestamp. When multiple events share the same source-destination pair, the edge weight increments and the first and last timestamps are preserved. Duplicates become weight, not separate edges. For flow events, a computer-to-computer edge stores protocol, source and destination ports, packet count, byte count, duration, and timestamp, with the same deduplication. Node types are inferred heuristically: node names ending with \texttt{\$} or containing \texttt{\$} before \texttt{@} are machine accounts; others are users. After all events are processed, the edge map is materialized into an igraph object.

Three graph variants are constructed per run: \emph{flow-only} (approximately 11,586 nodes, 131,562 edges), \emph{auth-only} (approximately 90,783 nodes, 409,100 edges), and \emph{combined} (approximately 91,589 nodes, 512,529 edges). Each graph is freed after scoring and evaluation.

\subsection{Feature Extraction}

We extract structural, temporal, and statistical features at the node, edge, and graph levels. All features are computed independently for each graph variant.

\textbf{Node features:}
For each node, we compute in-degree, out-degree, total degree, and fan-out ratio (out-degree divided by total degree). The fan-out ratio targets a specific lateral movement pattern: compromised hosts performing reconnaissance connect to many distinct destinations, while normal servers mostly receive inbound connections. Betweenness centrality is computed exactly for graphs under 5,000 nodes; for larger graphs, we use an approximation with a path length cutoff of 3 hops, trading some accuracy for tractability.

Temporal features capture timing patterns of outgoing connections. For each node, we compute mean and standard deviation of inter-arrival times, a burst score (the maximum fraction of outgoing edges falling within any 10\% sub-window of the node's active span), and total active duration. The intuition is that reconnaissance produces bursty, rapid-fire connections while persistent lateral movement shows longer, more regular intervals.

\textbf{Edge features:}
For each edge, we compute edge rarity (the reciprocal of edge weight), source out-degree, destination in-degree, and binary indicators for authentication signals: whether the authentication uses NTLM, whether the logon type is Network (remote access), and whether authentication succeeded. NTLM is particularly relevant because it is uncommon in normal traffic but frequently appears in pass-the-hash attacks.

For flow edges, we compute an unusual destination port indicator flagging ports associated with lateral movement (SSH-22, Telnet-23, SMB-445, RDP-3389, WinRM-5985/5986) or ephemeral ports above 49,152; protocol rarity (one minus the ratio of edges using that protocol to total flow edges); byte-per-packet ratio as a percentile rank; and duration z-score across all edges. Self-loops and user-to-user edges are excluded from scoring.

\textbf{Graph features:}
We compute density, average clustering coefficient, connected component count, node count, and edge count. These give a snapshot of each graph variant's structural complexity.

\subsection{Anomaly Scoring}

\subsubsection{Edge Scoring}
Edges are scored using a weighted combination of features, with the formula depending on edge type. The current weights are initial values chosen based on domain knowledge; we plan to tune them systematically in future work.

Authentication edges:
\begin{equation}
\begin{aligned}
s_{\text{auth}} ={}& 0.4\,\text{is\_ntlm} \\
&+ 0.3\,\text{is\_network\_logon} \\
&+ 0.3\,\text{rarity\_rank}
\end{aligned}
\end{equation}
NTLM receives the highest weight because pass-the-hash attacks are a common lateral movement technique. The network logon type indicates remote access rather than interactive sessions. Edge rarity (percentile-ranked) identifies uncommon source-destination pairs.

Flow edges:
\begin{equation}
\begin{aligned}
s_{\text{flow}} ={}& 0.4\,\text{rarity\_rank} \\
&+ 0.3\,\text{is\_unusual\_dst\_port} \\
&+ 0.3\,\text{protocol\_rarity\_rank}
\end{aligned}
\end{equation}
Rarity is weighted highest because uncommon connections are the strongest signal available from flow data. The unusual port flag encodes knowledge of lateral movement services, and protocol rarity measures how much a given protocol deviates from the normal distribution.

For the combined method, authentication edges receive a $1.5\times$ multiplier (\texttt{auth\_weight\_multiplier}) because auth signals are more informative but far fewer than flow edges. Without this multiplier, auth scores would be swamped. Self-loops and user-to-user edges are scored as 0.

\subsubsection{Unsupervised Baselines}

The manually-weighted formulas above are one of two scoring approaches we evaluate. In parallel, we run five unsupervised tabular anomaly detectors: One-Class SVM, Isolation Forest, Local
Outlier Factor, Elliptic Envelope, and PCA reconstruction error, on the same twelve-dimensional edge feature vector and the same $305$ red-team pairs (Table~\ref{tab:baselines_lanl}). Each detector is trained on benign edges only and scored on the full set, producing a valid comparison between the heuristic graph scoring and standard learned methods on identical ground truth.

\subsubsection{Path Enumeration and Scoring}
To capture multi-step lateral movement, we enumerate paths through the graph. Starting from each node, breadth-first search explores up to 4 hops, following only the top 10 outgoing edges per node ranked by score. Enumeration runs in parallel across 12 CPU workers.

Each path is scored as the mean of three aggregates of its edge scores: geometric mean (flags paths where every edge is somewhat suspicious), maximum edge score (flags paths with at least one highly anomalous hop), and arithmetic mean. The top 50 paths are retained. Edge scores then get a boost of 10\% of any traversing path's score (\texttt{path\_boost\_factor} $= 0.1$), feeding path-level signals back into edge-level detection.

\subsection{Threshold Optimization and Detection}

Anomalous edges are identified by thresholding final edge scores. Rather than fixing a percentile, we sweep across $[90, 95, 97, 99, 99.5, 99.9]$ and select the threshold maximizing F1 against known red-team pairs. This search operates in pair-space: after thresholding, anomalous source-destination pairs are compared against ground-truth pairs to compute recall, precision, and F1. We acknowledge that optimizing the threshold on the same labeled data used for evaluation is a limitation; the selected thresholds (95th percentile for auth-only, 97th for flow-only and combined) may not generalize to unseen attack patterns.

\section{Data and Experimental Setup}

\subsection{Datasets}

We evaluate on two public cybersecurity datasets.

\textbf{LANL Cybersecurity Dataset (LANL-2015).}
The primary dataset is the Los Alamos National Laboratory cybersecurity dataset from 2015~\cite{lanl2015}, covering 58 days of continuous network activity across an enterprise network. It contains approximately 1.6 billion events spanning 12,425 unique users and 17,684 unique computers, organized into gzipped CSV files: \texttt{auth.txt.gz} records authentication events (timestamp, source and destination users and computers, authentication type, logon type, orientation, and success/failure status); \texttt{flows.txt.gz} records network flow events (timestamp, duration, source and destination computers and ports, protocol, packet count, and byte count); and \texttt{redteam.txt.gz} provides ground-truth labels for 749 red team attack events spanning 308 unique source-destination host pairs. The red team events include lateral movement via SMB, SSH, RDP, credential reuse, and network reconnaissance, with 93.6\% of attacks originating from a single compromised host (C17693) used as a jump box.

To scope the data to a manageable size, we apply a time-windowing strategy. For each red team event, we define a window of $\pm$3,600 seconds around the event timestamp. Overlapping windows are merged, producing 25 non-overlapping intervals that capture approximately 104 million authentication events and 31.6 million flow events.

\textbf{DAPT2020 Dataset.}
The secondary dataset is DAPT2020~\cite{dapt2020}, which provides pre-extracted CICFlowMeter features from simulated APT attack traffic. It contains 86,690 flow records with 77 numerical features including packet length statistics, inter-arrival times, flag counts, and segment size averages, along with activity and stage labels. Lateral movement samples are identified by filtering the \texttt{Stage} column for entries containing ``lateral movement.'' We use this dataset to benchmark our graph-based approach against classical anomaly detection methods that operate on tabular features.

\subsection{Evaluation Metrics}

We report four metrics, all computed in pair-space: \emph{recall} (fraction of red-team pairs detected), \emph{false positive rate} (fraction of non-red-team pairs flagged), \emph{F1 score} (harmonic mean of precision and recall), and \emph{AUC} (computed using scikit-learn's \texttt{roc\_auc\_score} over all valid edges, with labels from the red-team pair set). AUC is threshold-independent: it measures whether the scoring ranks attack edges above normal ones regardless of where you set the cutoff. 308 red-team pairs among hundreds of thousands of total edges means precision will always be low, and the precision-recall tradeoff is steep.

\subsection{Baseline Methods}

We compare against two classical anomaly detection methods on DAPT2020: One-Class SVM (RBF kernel, $\gamma{=}\text{``scale''}$, $\nu{=}0.1$) and Isolation Forest (100 estimators, contamination$= 0.05$). Both are trained on normal-only samples and tested on data containing both normal and lateral movement traffic, using the 77 CICFlowMeter features. An important caveat: these baselines run on DAPT2020 while our graph methods run on LANL-2015, so the numbers are not directly comparable. Running all methods on both datasets is planned future work.

\subsection{Implementation Details}

The pipeline is implemented in Python using igraph for graph operations and scikit-learn for baselines and metrics. All hyperparameters live in a JSON config file (\texttt{pipeline\_config.json}). The scoring weights (Equations 1--2), the auth weight multiplier ($1.5\times$), the path boost factor ($0.1$), the hop limit (4), and the top-$k$ path count (50) are initial values that have not been tuned. Our initial results use these defaults. We intend to do proper hyperparameter optimization (grid search, Bayesian optimization, etc.) in the future. Due to the size of the dataset, we stream the data instead of extracting the entire dataset in memory. End-to-end execution for all three graph variants plus baselines takes approximately 4 hours on a 12-core machine.

\section{Results}
\begin{itemize}
\item Detection rate: percentage of red-team lateral movement events detected by each method (flow-only, auth-only, combined)
\item False positive rate: percentage of normal traffic flagged as suspicious by each method
\item Detection latency: time from attack event occurrence to alert generation
\item Throughput: number of log records processed per second by the pipeline
\item System overhead: CPU and memory usage of graph construction and scoring components
\item Comparison: combined method vs flow-only baseline vs auth-only baseline, with discussion of cases where each method succeeds or fails
\item Feature analysis: which graph features contribute most to detection (structural vs temporal vs statistical)
\item Key findings: TCP dominance in attack traffic (99.82\% TCP in red-team flows), kill-chain progression patterns (fan-out ratio increasing from 2.50 in reconnaissance to 6.0 in lateral movement), temporal separation (red-team median inter-arrival time 214s vs 6s for normal users)
\end{itemize}

\subsection{Tabular Baselines on LANL-2015}
\label{sec:baselines_lanl}

Table~\ref{tab:methods_lanl} reports all eight methods evaluated on
the same LANL-2015 run: three graph-based variants using the
manually-weighted scoring formulas, and five unsupervised tabular
detectors using \texttt{scikit-learn} defaults trained on benign
edges only. All eight methods are scored against the same 305
red-team source-destination pairs.

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\hline
Method & Recall & FPR & F1 & AUC \\
\hline
\multicolumn{5}{l}{\textit{Graph-based (pair space)}} \\
combined              & 0.662 & 0.020 & 0.039 & \textbf{0.954} \\
auth\_only            & 0.747 & 0.030 & 0.036 & 0.951 \\
flow\_only            & 0.013 & 0.030 & 0.002 & 0.579 \\
\hline
\multicolumn{5}{l}{\textit{Unsupervised tabular baselines (edge space)}} \\
Isolation Forest      & 0.741 & 0.050 & 0.024 & 0.940 \\
Elliptic Envelope     & 0.000 & 0.050 & 0.000 & 0.774 \\
One-Class SVM         & 0.492 & 0.098 & 0.008 & 0.706 \\
PCA Reconstruction    & 0.033 & 0.050 & 0.001 & 0.667 \\
Local Outlier Factor  & 0.207 & 0.045 & 0.008 & 0.493 \\
\hline
\end{tabular}
\caption{All eight methods on LANL-2015 (run \texttt{20260502\_165755}, 364{,}802-edge feature matrix, 305 red-team pairs). Graph methods evaluate threshold-dependent metrics in pair space (deduplicated source-destination pairs); tabular baselines evaluate in edge space. AUC is comparable across both spaces.}
\label{tab:methods_lanl}
\end{table}

The combined graph method is strongest overall (AUC 0.954). Among
baselines, Isolation Forest leads (AUC 0.940), within 0.015 of
combined and within 0.011 of auth-only. LOF underperforms on the
full data because the feature matrix is dominated by binary
indicators, which saturate the local-density signal LOF relies on.
The graph method's edge over the best baseline is small in AUC but
larger in false-positive rate


\section{Discussion and Limitations}
\begin{itemize}
\item Interpretation of expected findings: does combining log sources improve detection as hypothesized?
\item Limitations of datasets: LANL represents a single cloud VPC network over 58 days with red-team attacks that may not capture all real-world techniques; DAPT2020 is a simulated environment that may not reflect full cloud VPC network complexity
\item Limited attack diversity: red-team events focus on specific lateral movement techniques (SMB, SSH, credential reuse), does not include living-off-the-land or fileless approaches
\item Comparison with related work results: how does our approach compare to Hopper, Euler, Kitsune in terms of detection and false positive rates (acknowledging different datasets)
\item Real-world deployment considerations: scalability to larger cloud VPC networks, handling encrypted traffic, integration with existing SIEM systems, log storage costs
\end{itemize}

\section{Conclusion}
\begin{itemize}
\item Summary: presented a graph-based approach combining network flow logs and authentication logs for lateral movement detection in cloud VPC networks, evaluated on LANL-Dataset-2015 and DAPT2020
\item Key takeaways: combining log sources in a unified graph improves detection over single-source baselines; structural and temporal graph features capture lateral movement patterns that individual log analysis misses; TCP dominance, rising fan-out, and anomalous timing provide strong detection signals
\item Future work: test on additional real-world cloud VPC datasets, integrate with cloud-native security services, explore machine learning classifiers on graph features, investigate encrypted traffic analysis, extend to live detection deployments
\end{itemize}

{\footnotesize \bibliographystyle{acm}
\bibliography{sample}}

\end{document}
