\section{Methodology}

We ask two questions: does combining flow and authentication logs in a graph improve detection over single-source methods, and how does this approach compare to classical anomaly detection on tabular features?

\subsection{Design Rationale}

Our pipeline design draws on established knowledge about how lateral movement works in practice. Attackers moving through a network rely on TCP-based services --- SMB (port 445), SSH (port 22), RDP (port 3389) --- because these protocols provide the reliable, ordered delivery needed for remote command execution and file transfer. During reconnaissance, a compromised host typically connects to many destinations in a short period (high fan-out), then concentrates on specific targets once a path is found. The timing of attacker activity also tends to differ from normal users: lateral movement involves deliberate, spaced-out connections rather than the rapid, routine traffic patterns of legitimate users.

These patterns lead to three design decisions. We include protocol and port features in edge scoring to reflect the service-specific nature of lateral movement. We compute fan-out and degree features at the node level to detect the scanning-to-targeting transition. We also extract temporal features (inter-arrival times, burst scores, active duration) to separate deliberate attacker behavior from routine traffic.

\subsection{Datasets}

We evaluate on two public cybersecurity datasets:

\textbf{LANL Cybersecurity Dataset (LANL-2015).}
The primary dataset is the Los Alamos National Laboratory cybersecurity dataset from 2015~\cite{lanl2015}, covering 58 days of continuous network activity across an enterprise network. It contains approximately 1.6 billion events spanning 12,425 unique users and 17,684 unique computers, organized into gzipped CSV files: \texttt{auth.txt.gz} records authentication events (timestamp, source and destination users and computers, authentication type, logon type, orientation, and success/failure status); \texttt{flows.txt.gz} records network flow events (timestamp, duration, source and destination computers and ports, protocol, packet count, and byte count); and \texttt{redteam.txt.gz} provides ground-truth labels for 749 red team attack events spanning 308 unique source-destination host pairs. The red team events include lateral movement via SMB, SSH, RDP, credential reuse, and network reconnaissance, with 93.6\% of attacks originating from a single compromised host (C17693) used as a jump box.

To scope the data to tractable and relevant periods, we apply a time-windowing strategy. For each red team event, we define a window spanning $\pm$3,600 seconds around the event timestamp. Overlapping windows are merged, producing 25 non-overlapping intervals that capture approximately 104 million authentication events and 31.6 million flow events.

\textbf{DAPT2020 Dataset.}
The secondary dataset is DAPT2020~\cite{dapt2020}, which provides pre-extracted CICFlowMeter features from simulated APT attack traffic. It contains 86,690 flow records with 77 numerical features including packet length statistics, inter-arrival times, flag counts, and segment size averages, along with activity and stage labels. Lateral movement samples are identified by filtering the \texttt{Stage} column for entries containing ``lateral movement.'' We use this dataset to benchmark our graph-based approach against classical anomaly detection methods that operate on tabular features.

\subsection{Graph Construction}

We model the network as a directed multigraph using the igraph library, where nodes are entities (computers and users) and edges are interactions (authentication events and network flows). The graph is built incrementally by a \texttt{StreamingGraphBuilder} that reads log files line-by-line through time windows derived from the red-team ground truth. Events outside all windows are discarded immediately, keeping memory proportional to the active window set rather than the full dataset. This lets the pipeline process over 135 million events in approximately 4 GB of RAM.

For authentication events, the builder creates a computer-to-computer edge (source computer to destination computer) and optionally a user-to-user edge. Each edge stores authentication type, logon type, orientation, success status, and timestamp. When multiple events share the same source-destination pair, the edge weight is incremented and the first and last timestamps are preserved, producing a deduplicated graph with frequency information. For flow events, a computer-to-computer edge stores protocol, source and destination ports, packet count, byte count, duration, and timestamp. Duplicate pairs are collapsed with incremented weight. Node types are inferred heuristically: node names ending with \texttt{\$} or containing \texttt{\$} before \texttt{@} are machine accounts; others are users. After all events are processed, the edge map is materialized into an igraph object.

Three graph variants are constructed per run: \emph{flow-only} (approximately 11,586 nodes, 131,562 edges), \emph{auth-only} (approximately 90,783 nodes, 409,100 edges), and \emph{combined} (approximately 91,589 nodes, 512,529 edges). Each graph is freed after scoring and evaluation.

\subsection{Feature Extraction}

We extract structural, temporal, and statistical features at the node, edge, and graph levels. All features are computed independently for each graph variant.

\textbf{Node features.}
For each node, we compute in-degree, out-degree, total degree, and fan-out ratio (out-degree divided by total degree). The fan-out ratio targets a specific lateral movement pattern: compromised hosts performing reconnaissance connect to many distinct destinations, while normal servers mostly receive inbound connections. Betweenness centrality is computed exactly for graphs under 5,000 nodes; for larger graphs, we use an approximation with a path length cutoff of 3 hops, trading some accuracy for tractability.

Temporal features capture timing patterns of outgoing connections. For each node, we compute mean and standard deviation of inter-arrival times, a burst score (the maximum fraction of outgoing edges falling within any 10\% sub-window of the node's active span), and total active duration. The intuition is that reconnaissance produces bursty, rapid-fire connections while persistent lateral movement shows longer, more regular intervals.

\textbf{Edge features.}
For each edge, we compute edge rarity (the reciprocal of edge weight), source out-degree, destination in-degree, and binary indicators for authentication signals: whether the authentication uses NTLM, whether the logon type is Network (remote access), and whether authentication succeeded. NTLM is particularly relevant because it is uncommon in normal traffic but frequently appears in pass-the-hash attacks.

For flow edges, we compute an unusual destination port indicator flagging ports associated with lateral movement (SSH-22, Telnet-23, SMB-445, RDP-3389, WinRM-5985/5986) or ephemeral ports above 49,152; protocol rarity (one minus the ratio of edges using that protocol to total flow edges); byte-per-packet ratio as a percentile rank; and duration z-score across all edges. Self-loops and user-to-user edges are excluded from scoring.

\textbf{Graph features.}
We compute density, average clustering coefficient, connected component count, node count, and edge count. These provide context for interpreting each graph variant's structural complexity.

\subsection{Anomaly Scoring}

\subsubsection{Edge Scoring}
Edges are scored using a weighted combination of features, with the formula depending on edge type. The current weights are initial values chosen based on domain knowledge; we plan to tune them systematically in future work.

Authentication edges:
\begin{equation}
s_{\text{auth}} = 0.4 \cdot \text{is\_ntlm} + 0.3 \cdot \text{is\_network\_logon} + 0.3 \cdot \text{rarity\_rank}
\end{equation}
NTLM receives the highest weight because pass-the-hash attacks are a common lateral movement technique. Network logon type indicates remote access rather than interactive sessions. Edge rarity (percentile-ranked) identifies uncommon source-destination pairs.

Flow edges:
\begin{equation}
s_{\text{flow}} = 0.4 \cdot \text{rarity\_rank} + 0.3 \cdot \text{is\_unusual\_dst\_port} + 0.3 \cdot \text{protocol\_rarity\_rank}
\end{equation}
Rarity is weighted highest because uncommon connections are the strongest signal available from flow data. The unusual port flag encodes knowledge of lateral movement services, and protocol rarity measures how much a given protocol deviates from the normal distribution.

For the combined method, authentication edges receive a $1.5\times$ multiplier (\texttt{auth\_weight\_multiplier}) so that the more informative auth signals are not drowned out by the larger number of flow edges. Self-loops and user-to-user edges are scored as 0.

\subsubsection{Path Enumeration and Scoring}
To capture multi-step lateral movement, we enumerate paths through the graph. Starting from each node, breadth-first search explores up to 4 hops, following only the top 10 outgoing edges per node ranked by score. Enumeration runs in parallel across 12 CPU workers.

Each path is scored as the mean of three aggregates of its edge scores: geometric mean (flags paths where every edge is somewhat suspicious), maximum edge score (flags paths with at least one highly anomalous hop), and arithmetic mean. The top 50 paths are retained. Edge scores then receive a boost equal to 10\% of any traversing path's score (\texttt{path\_boost\_factor} $= 0.1$), so edges that participate in high-scoring paths get additional signal.

\subsection{Threshold Optimization and Detection}

Anomalous edges are identified by thresholding final edge scores. Rather than fixing a percentile, we sweep across $[90, 95, 97, 99, 99.5, 99.9]$ and select the threshold maximizing F1 against known red-team pairs. This search operates in pair-space: after thresholding, anomalous source-destination pairs are compared against ground-truth pairs to compute recall, precision, and F1. We acknowledge that optimizing the threshold on the same labeled data used for evaluation is a limitation --- the selected thresholds (95th percentile for auth-only, 97th for flow-only and combined) may not generalize to unseen attack patterns.

\subsection{Evaluation Metrics}

We report four metrics in pair-space: \emph{recall} (fraction of red-team pairs detected), \emph{false positive rate} (fraction of non-red-team pairs flagged), \emph{F1 score} (harmonic mean of precision and recall), and \emph{AUC} (computed using scikit-learn's \texttt{roc\_auc\_score} over all valid edges, with labels from the red-team pair set). AUC provides a threshold-independent measure of how well the scoring separates attack edges from normal ones. Pair-space evaluation is inherently difficult: 308 red-team pairs among hundreds of thousands of total edges makes the precision-recall tradeoff steep.

\subsection{Baseline Methods}

We compare against two classical anomaly detection methods on DAPT2020: One-Class SVM (RBF kernel, $\gamma{=}\text{``scale''}$, $\nu{=}0.1$) and Isolation Forest (100 estimators, contamination$= 0.05$). Both are trained on normal-only samples and tested on data containing both normal and lateral movement traffic, using the 77 CICFlowMeter features. A methodological caveat: these baselines run on DAPT2020 while our graph methods run on LANL-2015, so direct numerical comparison is not meaningful. Running both sets of methods on both datasets is planned future work.

\subsection{Implementation Details}

The pipeline is implemented in Python using igraph for graph operations and scikit-learn for baselines and metrics. All hyperparameters live in a JSON config file (\texttt{pipeline\_config.json}). Several of these --- the scoring weights (Equations 1--2), the auth weight multiplier ($1.5\times$), the path boost factor ($0.1$), the hop limit (4), and the top-$k$ path count (50) --- are initial values that have not been systematically tuned. We report results with these defaults and leave proper hyperparameter optimization (grid search, Bayesian optimization) to future work. Streaming construction processes the full LANL-2015 dataset (135M+ events in the time windows) in a single pass using approximately 4 GB of RAM. End-to-end execution for all three graph variants plus baselines takes approximately 4 hours on a 12-core machine.
